Controlled Experiments for Word Embeddings

### 2015-09-07 article/main.tex
hopefully final round of changes

\begin{section}{Introduction}
Word embeddings represent words from the vocabulary as dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense
vectors of dimension much smaller than the vocabulary size are
constructed such that they carry syntactic and semantic information.
Irrespective of the technique chosen, word embeddings are
typically derived from word co-occurrences.  More specifically, in a
machine-learning setting, word embeddings are typically trained by
scanning a short window over all the text in a corpus.  This process can
be seen as sampling word co-occurrence distributions, where it is
recalled that the co-occurrence distribution of a target word $w$
denotes the conditional probability $\p(w'|w)$ that a word $w'$ occurs in
its context, i.e. given that $w$ occurred.  Most applications of word
embeddings explore not the word vectors themselves, but relations
between them to solve, for example, similarity and word relation tasks
\cite{vecchi-baroni-zamparelli2011}.  For these tasks, it was found that using normalised word
vectors improves performance.  Word vector length is therefore typically
ignored.

@@

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}, who also based their work on word co-occurrences.
They define the contextual distinctiveness of a target word $w$ as the
Kullback-Leibler divergence
between the global word distribution $\p(w')$, and the
co-occurrence distribution $\p(w' | w)$ of the target word $w$.
So defined, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More specifically, if a
word $w$ tends to appear in a wide variety of contexts, the distribution
of the words it occurs with, i.e. the posterior distribution $\p(w' | w)$,
tends to diverge less from the prior distribution $\p(w')$, and its
contextual distinctiveness is accordingly small.  If, on the other hand,
$w$ typically appears in only a small number of different contexts, the
posterior distribution tends to diverge more from the prior, and a
larger value of contextual distinctiveness results.

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  In that paper, which is also based
on word co-occurrence, the authors study adjective-noun composites.
For each item in the vocabulary, they
recorded their co-occurrence within the same sentence with 
 frequent content words (nouns, adjectives or verbs) in the
corpus, and constructed word embeddings via 
singular value decomposition of the co-occurrence matrix.  The authors considered several models for
constructing vectors of unattested adjective-noun composites, the two
simplest being adding and component-wise multiplying the adjective and
noun vectors.  They hypothesized that the length of the vectors thus
constructed can be used to distinguish acceptable and semantically
deviant adjective-noun composites.  Using a few hundred adjective-noun
composites selected by humans for evaluation, they found that deviant
composites have a shorter vector than acceptable ones, in accordance
with their expectation.

In the word2vec similarity and word relationship tasks, normalised word vectors are used.
Word vector length is thus disregarded.
In the word2vec forum\footnote{\googlegroup}, it has been observed that normalisation improves performance on these tasks since word vector length is related to word frequency.
It remained unclear, however, whether word frequency was \textit{directly} related to word vector length or rather via correlated factors such as word significance.
Our experimental approach demonstrates that a direct relationship indeed exists.
It is shown moreover that the normalisation of the word vectors also prevents word co-occurrence noise from affecting the performance of the word similarity and word relationship tasks.
On the other hand, recent theoretical work \cite{Arora2015} has approached the problem of explaining the so-called ``compositionality'' property exhibited by some word embeddings.
In that work, unnormalised vectors are used in their model of the word relationship task.
It is hoped that experimental approaches such as those described here might enable theoretical investigations to describe the role of the word vector length in the word relationship tasks.
\end{section}

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia data dump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word occurrence counts are summarised 
in Table \ref{word-occurrences-table}.  This base corpus is then
modified as described in Sections \ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

@@

For simplicity only the word2vec CBOW word embedding with a single set
of hyperparameters is considered.  Specifically, a CBOW model with a
hidden layer of size 100 is trained using negative sampling with 5
negative samples, a window size of $10$, a minimum word occurrence of
$128$, and $10$ passes through the corpus.  Sub-sampling was not used so
that the influence of word frequency could be more clearly discerned.
Similar experimental results were obtained using hierarchical softmax,
but these are omitted for succinctness.  The relatively high
low-frequency cut-off is chosen to ensure that word vectors, in all but
degenerate cases, receive a sufficient number of gradient updates to be
meaningful.  This frequency cut-off results in a vocabulary of $81117$
words (only unigrams were considered).

@@

The replacement procedure of Section \ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
number of occurrences sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum occurrence count of any token in
the corpus.  Figure \ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Words chosen for the word frequency variation experiment, along with their occurrence counts in the unmodified corpus. }
	\label{word-frequency-counts}
\end{table}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus so
that its global occurrence rate is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the global word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure \ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher word occurrence rate of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of inserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure \ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a occurrence rate of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

@@

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and
so a limited number of samples is required to capture their
co-occurrence distributions.  We therefore expect that the vectors
representing the tokens derived from any of these words can be properly
learned from the modified corpus, even though their combined number of
occurrences equals the term frequency of that word in the original
corpus.  The red blocks in Figure \ref{word-frequency-experiment-heatmap}
clearly demonstrate that the direction of these vectors is constant for
each family of tokens.  In other words, the direction of these vectors
is only determined by the original word and is independent of how often
that words appear, provided sufficient samples have been processed for
the word vector to settle in its equilibrium direction.

@@

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Figure \ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the number of occurrences of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e. the plotted lines rarely cross.

@@

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs. frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section \ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are
          included for completeness.  }
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

\begin{section}{Varying co-occurrence noise}\label{CNVE}
This second experiment is complementary to the first.  Whereas in the
first experiment we studied the effect of word frequency on word vectors
for fixed co-occurrence, we next study the effect of co-occurrence for
fixed frequency.

\subsection{Generating noise}
We do so again in a controlled manner by adding noise to the
co-occurrence distribution.  We take the noise distribution to be the
global word distribution, so that noise can be simply added to the
co-occurrence distribution of a word by interspersing occurrences of
that word uniformly at random throughout the corpus.  A word that is
consistently used in a distinctive context in the unmodified corpus thus
appears in the modified corpus also in completely unrelated contexts.
As in Section \ref{WFVE}, we choose a small number of words from the
unmodified corpus for this experiment.  Table
\ref{cooccurrence-noise-words} lists the words chosen, along with their
frequencies in the corpus.

@@

\begin{subsection}{Experimental results}
Figure \ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity
of pairs of vectors representing some of the tokens used in this
experiment.  The figure clearly demonstrates that vectors representing
tokens derived from the same word all share a common direction.  This
brings us to the conclusion that co-occurrence noise has no discernible effect on
the direction of word vectors.

%While the word2vec neural network
%samples co-occurrence noise, the direction of the vector representing
%the target word changes, but these changes cancel out, so that the
%direction remains fixed on average.

The left panel in Figure \ref{fig:co-occurrence-noise-graph}, on the other hand, reveals that vector length varies more or less
linearly with the proportion $r$ of noise in the co-occurrence
distribution.  This figure motivates our interpretation of vector
length, within a sufficiently narrow frequency band, as a measure of the
absence of co-occurrence noise, or put differently, of the extent to
which a word carries the meaning of a distinctive context.

%When read from right to left, this figure can be reinterpreted and
%related to the corresponding figure in
%Figure \ref{fig:word-frequency-experiment-graph} as follows.
If the average effect of co-occurrence noise were neglible, then the horizontal axis of Figure \ref{fig:co-occurrence-noise-graph} could be reinterpreted as
denoting word frequency by reading the graph from right to left.
Thus a similar picture to Figure \ref{fig:word-frequency-experiment-graph} would emerge, namely that the
word vector length increases more or less linearly with word frequency.
This can not, however, be the complete picture, for the vectors
represented by the second synaptic layer, syn1neg, show a different
behavior in the two figures.  In particular, the length of these vectors
is seen from the right panel of Figure \ref{fig:co-occurrence-noise-graph}
to be completely insensitive to noise, i.e. frequency in our
reinterpretation, whereas the length of those shown in the right panel
of Figure \ref{fig:word-frequency-experiment-graph} does depend on
frequency.

In this experiment, the introduced tokens are intermediate points on a path in the space of distributions over the vocabulary, beginning at the co-occurrence distribution of the word and ending at the global word frequency distribution, i.e. the co-occurrence distribution of \word{VOID}.
Thus it is clear a priori that the word vectors for the introduced tokens are intermediate steps on a path from the word vector of the original word to that of \word{VOID} (recall from Subsection \ref{WFVE-results} that word vector of \word{VOID} is approximately independent of its frequency).
Thus, while we do indeed observe from Figure \ref{fig:co-occurrence-noise-heatmap} that the word vectors for the tokens introduced for a word all share a common direction, this is only true because the word vector for \word{VOID} is close to the origin.

@@

\begin{sidewaysfigure*}
	\includegraphics[scale=0.6]{cooccurrence-noise-graph}
	\caption{ Vector length vs. proportion of co-occurrence noise
          for words chosen for this experiment.  For each word, tokens
          of equal frequency but with increasing proportion of
          co-occurrence noise were inserted into the corpus, as
          described in section \ref{CNVE}.  The word vectors are
          obtained from the first synaptic layer, syn0.  The second
          layer, syn1neg, is included for completeness.  }
	\label{fig:co-occurrence-noise-graph}
\end{sidewaysfigure*}

@@

\begin{section}{Discussion}\label{future-directions}
Our principle contribution has been to demonstrate that controlled experiments can be used to gain insight into a word embedding.
These experiments can be carried out for any word embedding (or indeed language model), because they are achieved via modification of the training corpus.
Indeed, the experiments require no knowledge or modification of the model implementation.
It would naturally be of interest to perform these experiments for other word embeddings other than word2vec CBOW, such as skipgrams and GloVe, as well as for different hyperparameters settings.

@@

Further afield, it would be interesting to perform analogs of these experiments in other contexts where models are trained from co-occurrence data.
For instance, in the context of e-Commerce, product purchase frequency and product co-occurrence noise can be varied by modifying the product-product co-occurrence matrix.

@@

The co-occurrence distribution of \word{VOID} is the global frequency distribution, and in this sense pure background noise.
Thus the word vector of \word{VOID} is a special point in the feature space.
Figure \ref{fig:word-frequency-experiment-graph} shows that this point is not at the origin of the feature space (i.e. is not the zero vector).
The origin, however, is implicitly the point of reference in the word2vec word similarity task.
This raises the question of whether improved performance on the similarity task could be achieved by transforming the feature space (or modifying the model) such that the representation of pure noise (i.e. the vector for \word{VOID}) is at the origin of the transformed feature space.

>>>>

\begin{section}{Introduction}
Word embeddings, or distributed representations of words, have been the subject
of much recent research in the natural language processing and machine learning
communities, demonstrating state-of-the-art performance on word similarity and
word analogy tasks, amongst others.
Word embeddings represent words from the vocabulary as dense, real-valued
vectors.  Instead of one-hot vectors that merely indicate the location
of a word in the vocabulary, dense
vectors of dimension much smaller than the vocabulary size are
constructed such that they carry syntactic and semantic information.
Irrespective of the technique chosen, word embeddings are
typically derived from word co-occurrences.  More specifically, in a
machine-learning setting, word embeddings are typically trained by
scanning a short window over all the text in a corpus.  This process can
be seen as sampling word co-occurrence distributions, where it is
recalled that the co-occurrence distribution of a target word $w$
denotes the conditional probability $\p(w'|w)$ that a word $w'$ occurs in
its context, i.e. given that $w$ occurred.  Most applications of word
embeddings explore not the word vectors themselves, but relations
between them to solve, for example, similarity and word relation tasks
\cite{vecchi-baroni-zamparelli2011}.  For these tasks, it was found that using normalised word
vectors improves performance.  Word vector length is therefore typically
ignored.

@@

\begin{section}{Related work}\label{related-work}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}.
They define the contextual distinctiveness of a target word $w$ as the
Kullback-Leibler divergence
between the unconditional word distribution $\p(w')$, and the
co-occurrence distribution $\p(w' | w)$ of the target word $w$.
So defined, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More specifically, if a word $w$ tends to appear in a wide variety of contexts, then its co-occurrence distribution tends to diverge less from unconditional word distribution, and its
contextual distinctiveness is accordingly small.  If, on the other hand,
$w$ typically appears in only a small number of different contexts, its co-occurrence distribution tends to diverge more from the unconditional word distribution, and a larger value of contextual distinctiveness results.
In order to reduce the computational burden, the authors considered only $690$ target words, randomly chosen from equally divided frequency bands.
Furthermore, the authors considered only co-occurrence with the $500$ most frequent content words.
This significant restriction, necessary in their case in order to reduce the variance of the estimates of their measure of contextual distinctiveness, does not apply in our study.

Our experimental finding that word vector length decreases with
co-occurrence noise is related to earlier work by Vecchi, Baroni, and
Zamparelli \cite{vecchi-baroni-zamparelli2011}, where a relation between
vector length and the ``semantic deviance'' of an adjective-noun
composite was studied empirically.  
In that paper, which is also based
on word co-occurrence, the authors study adjective-noun composites.
They built a vocabulary from the 8k most frequent nouns and 4k most frequent adjectives in a large general language corpus and added 22k adjective-noun composites.
For each item in the vocabulary, they recorded the co-occurrences with the top 10k most frequent content words (nouns, adjectives or verbs), and contructed word embeddings via singular value decomposition of the co-occurrence matrix \cite{landauer-dumais1997}.
The authors considered several models for constructing vectors of unattested adjective-noun composites, the two simplest being adding and component-wise multiplying the adjective and noun vectors.
They hypothesized that the length of the vectors thus constructed can be used to distinguish acceptable and semantically deviant adjective-noun composites.
Using a few hundred adjective-noun composites selected by humans for evaluation, they found that deviant composites have a shorter vector than acceptable ones, in accordance with their expectation.
In contrast to their work, our approach does not require extensive human evaluation.

Recent theoretical work \cite{Arora2015} has approached the problem of explaining the so-called ``compositionality'' property exhibited by some word embeddings.
In that work, unnormalised vectors are used in their model of the word relationship task.
It is hoped that experimental approaches such as those described here might enable theoretical investigations to describe the role of the word vector length in the word relationship tasks.
\end{section}

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia data dump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see\\ \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word frequencies are summarised 
in Table \ref{word-occurrences-table}.  This base corpus is then
modified as described in Sections \ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

@@

For simplicity only the word2vec CBOW word embedding with a single set
of hyperparameters is considered.  Specifically, a CBOW model with a
hidden layer of size 100 is trained using negative sampling with 5
negative samples, a window size of $10$, a minimum frequency of
$128$, and $10$ passes through the corpus.  Sub-sampling was not used so
that the influence of word frequency could be more clearly discerned.
Similar experimental results were obtained using hierarchical softmax,
but these are omitted for succinctness.  The relatively high
low-frequency cut-off is chosen to ensure that word vectors, in all but
degenerate cases, receive a sufficient number of gradient updates to be
meaningful.  This frequency cut-off results in a vocabulary of $81117$
words (only unigrams were considered).

@@

The replacement procedure of Section \ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
frequency sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum frequency of any token in
the corpus.  Figure \ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

\begin{table}
	\input{word-frequency-experiment-counts.tex}
	\caption{Words chosen for the word frequency variation experiment, along with their frequency in the unmodified corpus. }
	\label{word-frequency-counts}
\end{table}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}\label{WFVEmeaningless}
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus so
that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure \ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency of $0.05$ is used instead
for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of inserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure \ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a relative frequency of $0.05$.}
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

@@

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and
so a limited number of samples is required to capture their
co-occurrence distributions.  We therefore expect that the vectors
representing the tokens derived from any of these words can be properly
learned from the modified corpus, even though their combined frequency equals the frequency of that word in the original
corpus.  The red blocks in Figure \ref{word-frequency-experiment-heatmap}
clearly demonstrate that the direction of these vectors is constant for
each family of tokens.  In other words, the direction of these vectors
is only determined by the original word and is independent of how often
that words appear, provided sufficient samples have been processed for
the word vector to settle in its equilibrium direction.

@@

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Figure \ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the frequency of the last token is still above the
low-frequency cut-off.  Because all the points on a line share the same
co-occurrence distribution, the left panel in
Figure \ref{fig:word-frequency-experiment-graph} demonstrates conclusively
that length does indeed depend on frequency directly.  Moreover, this
relation is seen to be approximately linear for each word considered.
Notice also that the relative positions of the lengths of the word vectors associated
with the experiment words are roughly independent of the frequency band,
i.e. the plotted lines rarely cross.

@@

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs. frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section \ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are
          included for completeness.
	  Legend entries are ordered by the L2-norm at the lowest-frequency
  (i.e. left-most) sample point for the corresponding syn0 plot, descending.}
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

\begin{section}{Varying co-occurrence noise}\label{CNVE}
This second experiment is complementary to the first.  Whereas in the
first experiment we studied the effect of word frequency on word vectors
for fixed co-occurrence, we next study the effect of co-occurrence noise when the frequency is fixed.
As before, we do so in a controlled manner.

\subsection{Generating noise}
We take the noise distribution to be the (observed) unconditional word distribution.
Thus noise can be added to the co-occurrence distribution of a word by simply interspersing occurrences of
that word uniformly at random throughout the corpus.  A word that is
consistently used in a distinctive context in the unmodified corpus thus
appears in the modified corpus also in completely unrelated contexts.
As in Section \ref{WFVE}, we choose a small number of words from the
unmodified corpus for this experiment.  Table
\ref{cooccurrence-noise-words} lists the words chosen, along with their
frequencies in the corpus.

@@

\begin{subsection}{Experimental results}
Figure \ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity
of pairs of vectors representing some of the tokens used in this
experiment.  The figure clearly demonstrates that vectors representing
tokens derived from the same word all share a common direction.  Thus
co-occurrence noise has no discernible effect on the direction of word vectors.

On the other hand, for any experiment word, the introduced tokens mark
intermediate points on a path in the space of distributions over the
vocabulary, beginning at the co-occurrence distribution of the word and ending
at the unconditional word frequency distribution, i.e. the co-occurrence
distribution of \word{VOID}.
Thus, a priori, the word vectors for the introduced tokens are intermediate
steps on a path from the word vector of the original word to that of
\word{VOID} (recall from Subsection \ref{WFVE-results} that word vector of
\word{VOID} is approximately independent of its frequency).
Since the word vector for \word{VOID} is close to the origin (see Subsection \ref{WFVE-results}), it may therefore be concluded that this path is approximately linear.

The left panel in Figure \ref{fig:co-occurrence-noise-graph} reveals that vector length varies more or less
linearly with the proportion $r$ of noise in the co-occurrence
distribution.  This figure motivates an interpretation of vector
length, within a sufficiently narrow frequency band, as a measure of the
absence of co-occurrence noise, or put differently, of the extent to
which a word carries the meaning of a distinctive context.

It was observed that the average effect of noise occurrences on word vector direction was negligible.
If the average effect of noise occurrences on word vector \textit{length} were also negligible, then the phenomenon depicted in Figure \ref{fig:co-occurrence-noise-graph} could be accounted for by the word frequency variation effects described in Section \ref{WFVE}.
For then the horizontal axis of Figure \ref{fig:co-occurrence-noise-graph} could be reinterpreted as
denoting word frequency by reading the graph from right to left,
and a similar picture to Figure \ref{fig:word-frequency-experiment-graph} would emerge, namely that the
word vector length increases more or less linearly with word frequency.
This can not, however, be the complete picture, for the vectors
represented by the second synaptic layer, syn1neg, show a different
behavior in the two figures.  In particular, the length of these vectors
is seen from the right panel of Figure \ref{fig:co-occurrence-noise-graph}
to be completely insensitive to noise, i.e. frequency in our
reinterpretation, whereas the length of those shown in the right panel
of Figure \ref{fig:word-frequency-experiment-graph} does depend on
frequency.
We conclude that co-occurrence noise does indeed effect word vector length.
In particular, the results of the two experiments illustrate distinct phenomena.

@@

\begin{sidewaysfigure*}
	\includegraphics[scale=0.6]{cooccurrence-noise-graph}
	\caption{ Vector length vs. proportion of co-occurrence noise
          for words chosen for this experiment.  For each word, tokens
          of equal frequency but with increasing proportion of
          co-occurrence noise were inserted into the corpus, as
          described in section \ref{CNVE}.  The word vectors are
          obtained from the first synaptic layer, syn0.  The second
          layer, syn1neg, is included for completeness.  
	  Legend entries are ordered by the left-most value of the corresponding syn0 plot, descending.}
	\label{fig:co-occurrence-noise-graph}
\end{sidewaysfigure*}

@@

\begin{section}{Discussion}\label{future-directions}
Our principle contribution has been to demonstrate that controlled experiments can be used to gain insight into a word embedding.
These experiments can be carried out for any word embedding (or indeed language model), because they are achieved via modification of the training corpus.
Indeed, the experiments even do not require knowledge of the model implementation.
It would naturally be of interest to perform these experiments for other word embeddings other than word2vec CBOW, such as skipgrams and GloVe, as well as for different hyperparameters settings.

@@

@@

The co-occurrence distribution of \word{VOID} is the unconditional frequency distribution, and in this sense pure background noise.
Thus the word vector of \word{VOID} is a special point in the feature space.
Figure \ref{fig:word-frequency-experiment-graph} shows that this point is not at the origin of the feature space (i.e. is not the zero vector).
The origin, however, is implicitly the point of reference in the word2vec word similarity task.
This raises the question of whether improved performance on the similarity task could be achieved by transforming the feature space (or modifying the model) such that the representation of pure noise (i.e. the vector for \word{VOID}) is at the origin of the transformed feature space.
